Storm surge modelling in the Strait of Georgia

### 2014-12-03 RoughWork/manuscript.tex
Wordy corrections from Susan

More recently, modelling studies have focused on baroclinic features. \citet{masson2004modelling} used the Princeton Ocean Model (POM) in a regional mode to examine the seasonal variability of deep water properties in the Strait of Georgia noting a strong sensitivity to vertical mixing over the sills. This model employed a sigma or terrain-following vertical coordinate system leading to spurious diapycnal mixing since, numerically, the lateral mixing is computed on sigma levels which may intersect isopycnals in the presence of steep topography. A finite volume model on an unstructured grid by \citet{yang2010multi} involved a truncated domain with an open boundary in the northern Strait of Georgia, where under predictions in the M$_2$ tidal phase over most of their domain were suggested to be caused by an inaccurate representation of the tides at their northern boundary. \citet{sutherland2011model} employed the Regional Ocean Modelling System (ROMS) to study estuarine circulation in the Salish Sea with a model domain that extended outside the Strait of Juan de Fuca into the Pacific Ocean and up to the northern Strait of Georgia and noted that their semi-diurnal amplitudes were generally too low. An updated version of this model produced high temperatures in the Strait of Juan de Fuca, suggesting too little mixing within the Salish Sea \citep{giddings2014hindcasts}. Smaller domains have also been considered, notably, the passageways through the Discovery Islands north of the Strait of Georgia have been modelled baroclinically by \citet{foreman2012circulation} with the Finite Volume Coastal Ocean Model (FVCOM) pointing to difficulties in reproducing baroclinic currents on their grid.  Accurate reproduction of observed tidal amplitudes and phases and appropriate treatment of mixing is a recurring difficulty in modelling of this region.

@@

%A section about how we configured the model for SoG 
We have used the Nucleus for European Modelling of the Ocean (NEMO) framework in its regional configuration \citep{madec2012nemo} to develop an ocean model for the Strait of Georgia and Salish Sea. NEMO is a highly modularized tool used for studying ocean physics, ocean-ice interactions, and the bio-geochemical properties of the ocean. NEMO's ocean core solves the three-dimensional hydrostatic equations of motion for an incompressible fluid under the Boussinesq approximation on a structured computational grid. Although not used in the present work, NEMO's options for grid nesting and bio-geochemical coupling make it a useful tool for studying the complex physics and bio-geochemical interactions within the Strait of Georgia. This work focuses on validating the physical set up of the Salish Sea model, in particular, determining appropriate forcing and boundary conditions for accurate reproduction of tidal amplitudes and phases as well as storm surge elevations. Future work will include bio-geochemical coupling and data assimilation.

@@

In addition to the equations of motion, a prognostic equation for the sea surface height is solved at each time step. The inclusion of the sea surface height equation requires a fairly restrictive time step due to the presence of high speed surface gravity waves. As such, the split-explicit time stepping algorithm is employed, where the free surface and barotropic equations are solved with a smaller time step than that used for the other variables. The model time step and barotropic time step are 10 s and 2 s respectively. Small vertical grid spacing and large vertical velocities in the Boundary Pass and Haro Strait region impose a fairly restrictive time step through the vertical CFL condition.

@@

Sea surface height at the mouth of Juan de Fuca was set using values from the Tofino tide gauge which is outside of but close to the western boundary of the model domain.  A monthly climatology was produced using daily averages from 2000-2010, binning them by month, averaging and setting the yearly mean to zero.  For the storm surge simulations, hourly variations in sea surface height were used.  These values are the Tofino tide gauge values, de-tided and with the zero reset as for the climatology. The storm surge simulations also used the sea surface height anomaly from Port Hardy, forced at the northern boundary in Johnstone Strait. The tidal forcing and sea surface height was applied to the barotropic velocity forcing with the NEMO Flather scheme \citep{flather1994storm, madec2012nemo}. At the west, the baroclinic velocities at the boundary were set equal to the values inside the boundary (zero-gradient boundary conditions).  This scheme is not part of core NEMO 3.4, the version used here.  Zero gradient conditions were chosen because the baroclinic velocity at the mouth of Juan de Fuca is primarily estuarine and thus set by density variations between inside and outside the domain. At the north baroclinic velocities were set to zero.

@@

Freshwater runoff from each watershed was divided between the rivers in that watershed. The area drained by each river was estimated from Toporama maps by the Atlas of Canada and watershed maps available on the Washington State government website. The watersheds included in our model were Fraser (which represents approximately 44\% of the freshwater input into our domain), Skagit (12\%), East Vancouver Island (North and South) (12\%), Howe (7\%), Bute (7\%), Puget (6\%), Juan de Fuca (5\%), Jervis (4\%) and Toba (3\%). The monthly flow from each river was input as a point source at a depth of 3 metres at the model point closest to the mouth of each river. Incoming water was assumed to be fresh and at surface temperature. A total of 150 rivers were parametrized by this method and their locations are indicated by the blue dots in Figure \ref{fig:domain}.

@@

The model was initially evaluated qualitatively by comparing patterns of tidal amplitude and phase to results from \citet{foreman1995tidal}, focusing on the major M$_2$ and K$_1$ constituents. For example, the M$_2$ amphidrome around Victoria was reproduced as well as the monotonic increase in $K_1$ amplitude moving northwards along the Strait of Georgia. Once the model was reproducing observed tidal patterns, model results were quantitatively evaluated by comparing modelled harmonic constituents to measured harmonic constituents at tidal measuring stations throughout the domain. Comparisons were made using the complex difference ($D$), defined as \citep{foreman1995tidal}:

@@

\section{Storm Surge Hindcasts}\label{sec:storm}
A series of storm surge hindcasts is presented next. Storm surge simulations were started from rest at least three days prior to the event of interest to allow for adjustment of model currents. Stratification conditions from model spin-up at a time close to the event (within five days) were chosen to initialize the temperature and salinity fields. Boundary conditions include atmospheric forcing and fresh water input from rivers at the surface, sea surface height anomaly from observations at the open boundary, and tidal forcing with eight tidal constituents. Forcing the model with only eight tidal constituents leads to an error in modelled sea surface height due to the omission of the next leading order tidal constituents such as J$_1$. Comparisons between full tidal predictions and tidal predictions using eight constituents suggest that this error can approach 40 cm during times of high tidal range, and in particular, during the storm surge season Nov-Feb. As such, storm surge model predictions were corrected by accounting for this error. Additionally, since the sea surface height is calculated about a zero reference level, the modelled water level is determined by adding a long term mean sea level at the specified location to the modelled sea surface height.

Model results have been compared with observations by calculating both the total modelled water level and the modelled residual. The modelled residual is calculated as the difference in model output between a simulation with all forcing conditions and a simulation forced only with tides and rivers. Water level observations are taken from Fisheries and Oceans Canada (citation?) and are used to calculate the observed residual, defined as the observed water level minus tidal predictions. Tidal predictions are calculated using t\_tide \citep{pawlowicz2002classical}, a MATLAB tool that calculates tidal harmonics from a given time series, in this case, from the year prior surge event of interest. Additionally, several simulations with varying forcing conditions such as removing atmospheric forcing or removing the sea surface height anomaly at the open boundaries have been completed.

Several significant surge or wind events have been considered in detail, as well as an array of hindcasts to asses model ability in a statistical measure. These cases have been used to assess the model's skill in reproducing water level height and surge amplitude and timing during storm events. The assessment includes comparisons at four permanent tidal measuring stations; Point Atkinson, Victoria, Patricia Bay, and Campbell River (locations indicated in Figure \ref{fig:domain}).  In addition, the contribution to the surge due to several forcing conditions, such as atmospheric forcing and open boundary forcing, has been assessed. We have also examined the spatial variability of the surge as it propagates through the domain and have considered a case with a strong wind event but no significant observed surge.  Finally, comparisons between hindcasts using high resolution and low resolution atmospheric forcing have been completed.

@@

The model's ability to reproduce this event is demonstrated in Figure \ref{fig:feb2006} which compares observations to corrected model output at Point Atkinson, Victoria, Patricia Bay, and Campbell River. The timing of maximum water level matches well between the observations and the model at all four locations, the most significant contribution to high water being from the tides which are generally well-represented by the model. There is good agreement in the water level maximum at both Victoria and Patricia Bay, with a relative error less than 2\%. In fact, there is only a 4.2 cm difference between observed and modelled maximum at Victoria and a 6.9 cm difference at Patricia Bay, the model slightly under predicting in each case. The error at Point Atkinson and Campbell River is slightly more significant, reaching 2.5\% and 2.7\% respectively, close to 14 cm over predicted at both locations. It is surprising that Victoria and Patricia Bay agree so closely with observations because the M$_2$ tidal amplitudes in the model are known to be too low in this area. When considering the over predicted water levels at Point Atkinson and Campbell River, it is likely that one of the forcing factors is overestimated in this example.

A comparison between observed and modelled residuals is also displayed in the panels on the right. In general, there is a very good agreement between the observed and modelled residuals. The most noticeable difference is a 4-6 hour delay in the model's timing of the maximum residual except at Victoria where the maximum residual is in sync with observations. The errors in maximum residual range from 6.0 cm too high at Victoria to 15.9 cm too low at Campbell River. Both Point Atkinson and Patricia Bay experience a very good match with observed residuals with errors in the maximum less than 2 cm. Note that Point Atkinson and Campbell River realized the largest observed surges at 82 cm and 95 cm respectively. The surges at Victoria and Patricia Bay were observed to be 66 cm and 74 cm.

@@

The maximum water level and maximum residual are important factors in storm surge hindcasts but a better indicator of overall model performance is provided by a group of statistical measures, including the mean absolute error, the observed mean, the modelled mean, $\gamma^2$, and the Willmott skill score for both the water level and residuals, in Table \ref{tab:feb2006stat}. The statistics are calculated with hourly output over the three-day period displayed in Figure \ref{fig:feb2006}. To begin with, the mean absolute error in the water level ranges from 7.6 cm at Victoria to 13.1 cm at Campbell River, whereas, the mean absolute error in the residuals is smaller at all stations, however, this measure does not account for the large variability in water levels due to the tides. A statistic that does take into account this variability is $\gamma^2$ which suggests very good agreement in both the predicted water levels and residuals. Good agreement is also reflected in high Willmott skill scores at all locations. Note that the $\gamma^2$ measures and Willmott skill scores degrade slightly for the residuals, likely because the well-tuned tidal signal dominates the water level predictions.

@@

Next, an assessment of the importance of including the tide-surge interaction, which is important in shallow regions due to nonlinear effects and bottom friction, is presented for this hindcast. A simulation of the surge propagation without including any tidal forcing is compared with the modelled residual in Figure \ref{fig:tidesurge}. The most significant differences occur at Point Atkinson where the maximum of the surge-only case is 8.7 cm higher than maximum modelled residual, about 10\% of the total surge amplitude.  In contrast, the tide-surge interaction at Victoria is hardly noticeable. Indeed, a spatial map of the difference between the modelled residual and surge-only difference (not shown here) indicates that the tide-surge interaction is more pronounced after the San Juan and Gulf Islands. The passages between these islands are known for vigorous tidal mixing, which may have a nonlinear effect on the surge propagation. However, the effect is a relatively small overestimate of the surge elevation and no changes to the timing of the surge. So, in operational settings, performing a surge-only simulation and adding the predicted tides \textit{a posteriori} could be more computationally efficient with only a small decrease in the accuracy of the prediction.

@@

An additional storm surge hindcast for Nov 2006 is provided as another means of assessing model performance and is mentioned briefly here. On Nov 16, 2006 communities in the Strait of Georgia experienced a strong wind storm with speeds upwards of 20 m s$^{-1}$ near Point Atkinson resulting in observed water levels of 5.1 m and a maximum surge amplitude of 83.7 cm. The surge amplitude was lower at the other stations, reaching no more than 68 cm. The model performs reasonably well in this hindcast; Willmott skill scores are greater than 0.94 as outlined in Table \ref{tab:nov2006stat}.  The best match in maximum surge occurred at Campbell River with a 6.2 cm underestimate by the model. The timing of the maximum surge at Campbell River was identical with observations and too early at the other the stations.

@@

Next, an assessment of the factors that are most important in scenarios leading up to storm surges is presented for the Feb 2006 storm surge in Figure \ref{fig:factors}. The modelled residual from four simulations have been plotted. First, a simulation with all the forcing conditions, including tides, atmospheric forcing, rivers, and sea surface height anomaly, is shown in the dashed black curve. Next, a simulation with all forcing except the atmospheric forcing is presented in solid grey and a simulation with all forcing except the sea surface height anomaly is shown solid black. Finally, a simulation with only wind forcing, that is no anomaly and no inverse barometer effect due to pressure, is shown in the dashed grey curve. It is clear that the sea surface height anomaly contributes most significantly to the surge at each of these locations. When the sea surface height anomaly is not included as a forcing condition, the surge drops from a maximum 82 cm to 11 cm at Point Atkinson and its maximum occurs eight hours later.  A similar drop in the maximum surge and delay in timing of the maximum surge is observed at the other locations. The atmospheric model indicates southeasterly winds near Point Atkinson during the onset of the surge but after the peak surge has passed the winds shift to westerly. A strong westerly wind would cause an elevated sea surface height at Point Atkinson as water piles against the coast, explaining the delay in peak surge when only atmospheric forcing is used since the winds shift to westerly after the peak surge has passed. A simulation with only wind forcing produces a surge of 7.8 cm at Point Atkinson, a small fraction of the total observed surge, and only centimetres at the other locations. Note that this wind-induced surge occurs well after the peak surge occurred in both the observations and simulations with full forcing.

@@

These comparisons point to the importance of including, as a forcing condition at the open boundary, the surge entering the domain from the Pacific Ocean. Storms travelling over the Pacific Ocean towards the west coast of the northern United States and Canada can induce elevated sea levels along the coastline which then enter the Salish Sea system through the Strait of Juan de Fuca. In a model domain that does not include the Pacific Ocean, this effect must be included as a boundary condition. Previous simulations by \citet{murty1995storm} suggest that the inverse barometer effect and the surge entering the system through the Strait of Juan de Fuca contribute most significantly to storm surges in this domain.  A detailed assessment of the effect of the local atmospheric conditions and winds should be performed with high resolution atmospheric forcing data.

@@

\subsection{Spatial Extent of Surge}
On Dec 15, 2006, a major storm with wind gusts over 120 km hr$^{-1}$ caused significant damage to many coastal communities on the Pacific Northwest. Although the storm was very strong and the observed residual at Point Atkinson reached 84 cm, this event did not coincide with high tide and so no significant or reported flooding occurred in the Vancouver area \citep{forseth2006adaptation}. This event has been used to examine the propagation of the surge throughout the domain and spatial variations in the surge amplitude. First, an assessment of the accuracy of the hindcast is provided.

Point Atkinson observed a maximum surge amplitude of 84 cm, which is approximately 28 cm higher than the modelled residual, however the timing of the surge agrees almost exactly as displayed in Figure \ref{fig:dec2006}. The best agreement occurs at Victoria where the observed surge was 62 cm and the model returned 44 cm. At all of the locations, the model predicted the maximum surge to occur within 3 hours of the observed surge. Typically, the modelled surge was too early, in contrast to the later predictions in the Feb 4, 2006 case. Error and skill scores for the modelled water level and the modelled residual are provided in Table \ref{tab:dec2006stat}. All Willmott skill scores rank above 0.89 and the $\gamma^2$ measure also indicates good agreement between model and observations. The modelled residuals are on average too low and have a lower skill score than the Feb 2006 case. This could be due to a number of factors. First, the very large wind speeds observed during this storm are significantly underrepresented by the atmospheric model leading to lower wind stress and less piling of water against the coast. Second, the anomaly forcing at the open boundaries could be too low. The observed residual at Tofino, BC (a point near by but outside of the domain) is up to 10 cm lower than the observed residual at Neah Bay, WA (a point near by but inside of the domain). Since the storm surge is less sensitive to wind forcing at the surface than anomaly forcing at the open boundaries, the latter is likely the case.

@@

An evaluation of the spatial distribution of the surge in a simulation without atmospheric forcing is provided in Figure \ref{fig:spatial_sshonly}. In this case, the surge amplitude across the Strait of Georgia is uniform with no pronounced difference between the mainland side and Vancouver Island side. While the surge entering the domain from the Pacific Ocean is the leading order contribution, the winds may induce localized changes to the surge amplitude. Interestingly, the winds over the Strait of Juan de Fuca cause a decrease in surge amplitude in this region as the winds are acting to push the water towards the mainland.  Even without atmospheric forcing, there is still a significant difference in surge amplitude between the Strait of Georgia and the Strait of Juan de Fuca, perhaps because the Strait of Georgia is also affected by the surge entering the domain from Johnstone Strait at the northern boundary or the incoming surge wave is partially reflected from the mainland coast.

@@

\subsection{Strong Wind Event}
Next, a relatively strong wind event recorded at Point Atkinson on Nov 19, 2009 is considered. As recorded in the Environment Canada Climate Data archives \citep{ECClimateArchive}, the maximum hourly wind speed at Point Atkinson was 18.6 m s$^{-1}$ and the largest gust observed was 25.8 m s$^{-1}$ or 93 km hr$^{-1}$. In addition, large wind speeds over 15 m s$^{-1}$ persisted for over 9 hours. Yet, the surge associated with this event is rather small when compared to the previous two cases as shown in Figure \ref{fig:nov2009} where we have plotted the wind speed at Point Atkinson, sea surface height anomaly forcing at the open boundaries, and observed and modelled residuals at Point Atkinson. Two significant surge events are noted, the first occurring on Nov 19 with an observed amplitude of about 50 cm at Point Atkinson, and a second, larger surge of 62.5 cm on Nov 20. Surprisingly, the strong wind event of Nov 19 was associated with the smaller of the two surges. Neither of the surges occurred during a high tide and so no high water level events were recorded. Although not shown here, the other three locations also observed a maximum surge on Nov 20, which appears to be linked to a secondary high wind event and another surge entering the system at Tofino and Port Hardy.

>>>>

More recently, modelling studies have focused on baroclinic features. \citet{masson2004modelling} used the Princeton Ocean Model (POM) in a regional mode to examine the seasonal variability of deep water properties in the Strait of Georgia noting a strong sensitivity to vertical mixing over the sills. This model employed a sigma or terrain-following vertical coordinate system leading to spurious diapycnal mixing since, numerically, the lateral mixing is computed on sigma levels which may intersect isopycnals in the presence of steep topography. A finite volume model on an unstructured grid by \citet{yang2010multi} involved a truncated domain with an open boundary in the northern Strait of Georgia, where under predictions in the M$_2$ tidal phase over most of their domain were suggested to be caused by an inaccurate representation of the tides at their northern boundary. \citet{sutherland2011model} employed the Regional Ocean Modelling System (ROMS) to study estuarine circulation in the Salish Sea with a model domain that extended outside the Strait of Juan de Fuca into the Pacific Ocean and up to the northern Strait of Georgia and noted that their semi-diurnal amplitudes were generally too low. An updated version of this model produced high temperatures in the Strait of Juan de Fuca, suggesting too little mixing within the Salish Sea \citep{giddings2014hindcasts}. Smaller domains have also been considered; notably, the passageways through the Discovery Islands north of the Strait of Georgia have been modelled baroclinically by \citet{foreman2012circulation} with the Finite Volume Coastal Ocean Model (FVCOM) pointing to difficulties in reproducing baroclinic currents on their grid.  Accurate reproduction of observed tidal amplitudes and phases and appropriate treatment of mixing is a recurring difficulty in modelling this region.

@@

%A section about how we configured the model for SoG 
We have used the Nucleus for European Modelling of the Ocean (NEMO) framework in its regional configuration \citep{madec2012nemo} to develop an ocean model for the Strait of Georgia and Salish Sea. NEMO is a highly modularized tool used for studying ocean physics, ocean-ice interactions, and the bio-geochemical properties of the ocean. NEMO's ocean core solves the three-dimensional hydrostatic equations of motion for an incompressible fluid under the Boussinesq approximation on a structured computational grid. Although not used in the present work, NEMO's options for grid nesting and bio-geochemical coupling make it a useful tool for studying the complex physics and bio-geochemical interactions within the Strait of Georgia. This work focuses on validating the physical configuration of the Salish Sea model, in particular, determining appropriate forcing and boundary conditions for accurate reproduction of tidal amplitudes and phases as well as storm surge elevations. Future work will include bio-geochemical coupling and data assimilation.

@@

In addition to the equations of motion, a prognostic equation for the sea surface height is solved at each time step. The inclusion of the sea surface height equation requires a fairly restrictive time step due to the presence of high speed surface gravity waves. As such, the split-explicit time stepping algorithm is employed, where the free surface and barotropic equations are solved with a smaller time step than that used for the other variables. The model baroclinic time step and barotropic time step are 10 s and 2 s respectively. Small vertical grid spacing and large vertical velocities in the Boundary Pass and Haro Strait region impose a fairly restrictive baroclinic time step through the vertical Courant-Friedrichs-Lewy (CFL) condition.

@@

Sea surface height at the mouth of Juan de Fuca was set using values from the Tofino tide gauge which is outside of, but close to, the western boundary of the model domain.  A monthly climatology was produced using daily averages from 2000-2010, binning them by month, averaging and setting the yearly mean to zero.  For the storm surge simulations, hourly variations in sea surface height were used.  These values are the Tofino tide gauge values, de-tided and with the zero reset as for the climatology. The storm surge simulations also used the sea surface height anomaly from Port Hardy, forced at the northern boundary in Johnstone Strait. The tidal forcing and sea surface height was applied to the barotropic velocity forcing with the NEMO Flather scheme \citep{flather1994storm, madec2012nemo}. At the west, the baroclinic velocities at the boundary were set equal to the values inside the boundary (zero-gradient boundary conditions).  This scheme is not part of core NEMO 3.4, the version used here.  Zero gradient conditions were chosen because the baroclinic velocity at the mouth of Juan de Fuca is primarily estuarine and thus set by density variations between inside and outside the domain. At the north baroclinic velocities were set to zero.

@@

Freshwater runoff from each watershed was divided between the rivers in that watershed. The area drained by each river was estimated from Toporama maps by the Atlas of Canada and watershed maps available on the Washington State government website. The watersheds included in our model were Fraser (which represents approximately 44\% of the freshwater input into our domain), Skagit (12\%), East Vancouver Island (North and South) (12\%), Howe (7\%), Bute (7\%), Puget (6\%), Juan de Fuca (5\%), Jervis (4\%) and Toba (3\%). The monthly flow from each river was input as a point source from 0-3 metres at the model point closest to the mouth of each river. Incoming water was assumed to be fresh and at surface air temperature. A total of 150 rivers were parametrized by this method and their locations are indicated by the blue dots in Figure \ref{fig:domain}.

@@

The model was initially evaluated qualitatively by comparing patterns of tidal amplitude and phase to results from \citet{foreman1995tidal}, focusing on the major M$_2$ and K$_1$ constituents. For example, the M$_2$ amphidrome around Victoria was reproduced as well as the monotonic increase in $K_1$ amplitude moving northwards along the Strait of Georgia. As the model was reproducing observed tidal patterns, model results were quantitatively evaluated by comparing modelled harmonic constituents to measured harmonic constituents at tidal measuring stations throughout the domain. Comparisons were made using the complex difference ($D$), defined as \citep{foreman1995tidal}:

@@

\section{Storm Surge Hindcasts}\label{sec:storm}
A series of five storm surge hindcasts is presented next. Storm surge simulations were started from rest at least three days prior to the event of interest to allow for adjustment of model currents. Stratification conditions from model spin-up at a time close to the event (within five days) were chosen to initialize the temperature and salinity fields. Boundary conditions include atmospheric forcing and fresh water input from rivers at the surface, sea surface height anomaly from observations at the open boundary, and tidal forcing with eight tidal constituents. Forcing the model with only eight tidal constituents leads to an error in modelled sea surface height due to the omission of the next leading order tidal constituents such as J$_1$. Comparisons between full tidal predictions and tidal predictions using eight constituents suggest that this error can approach 40 cm during times of high tidal range, and in particular, during the storm surge season Nov-Feb. As such, storm surge model predictions were corrected by accounting for this error. Additionally, since the sea surface height is calculated about a zero reference level, the modelled water level is determined by adding a long term mean sea level at the specified location to the modelled sea surface height.

Model results have been compared with observations by calculating both the total modelled water level and the modelled residual. The modelled residual is calculated as the difference in model output between a simulation with all forcing conditions and a simulation forced only with tides and rivers. Water level observations are taken from Fisheries and Oceans Canada (citation?) and are used to calculate the observed residual, defined as the observed water level minus tidal predictions. Tidal predictions are calculated using t\_tide \citep{pawlowicz2002classical}, a MATLAB tool that calculates tidal harmonics from a given time series, in this case, from the year prior to the surge event of interest. Additionally, several simulations with varying forcing conditions such as removing atmospheric forcing or removing the sea surface height anomaly at the open boundaries have been completed.

Several significant surge or wind events have been considered in detail, as well as an array of hindcasts to assess model ability in a statistical measure. These cases have been used to assess the model's skill in reproducing water level height and surge amplitude and timing during storm events. The assessment includes comparisons at four permanent tidal measuring stations; Point Atkinson, Victoria, Patricia Bay, and Campbell River (locations indicated in Figure \ref{fig:domain}).  In addition, the contribution to the surge due to several forcing conditions, such as atmospheric forcing and open boundary forcing, has been assessed. We have also examined the spatial variability of the surge as it propagates through the domain and have considered a case with a strong wind event but no significant observed surge.  Finally, comparisons between hindcasts using high resolution and low resolution atmospheric forcing have been completed.

@@

The model's ability to reproduce this event is demonstrated in Figure \ref{fig:feb2006} which compares observations to corrected model output at Point Atkinson, Victoria, Patricia Bay, and Campbell River. The timing of maximum water level matches well between the observations and the model at all four locations, the most significant contribution to high water being from the tides which are generally well-represented by the model. There is good agreement in the water level maximum at both Victoria and Patricia Bay, with a relative error less than 2\%. In fact, there is only a 4.2 cm difference between observed and modelled maximum at Victoria and a 6.9 cm difference at Patricia Bay, the model slightly under predicting in each case. The error at Point Atkinson and Campbell River is slightly more significant, reaching 2.5\% and 2.7\% respectively, close to 14~cm over predicted at both locations. It is surprising that Victoria and Patricia Bay agree so closely with observations because the M$_2$ tidal amplitudes in the model are known to be too low in this area. When considering the over predicted water levels at Point Atkinson and Campbell River, it is likely that one of the forcing factors is overestimated in this example.

In general, there is a very good agreement between the observed and modelled residuals \ref{fig:feb2006}. The most noticeable difference is a 4-6 hour delay in the model's timing of the maximum residual except at Victoria where the maximum residual is in sync with observations. The errors in maximum residual range from 6.0 cm too high at Victoria to 15.9 cm too low at Campbell River. Both Point Atkinson and Patricia Bay experience a very good match with observed residuals with errors in the maximum less than 2 cm.

@@

The maximum water level and maximum residual are important factors in storm surge hindcasts but a better indicator of overall model performance is provided by a group of statistical measures, including the mean absolute error, the observed mean, the modelled mean, $\gamma^2$, and the Willmott skill score for both the water level and residuals (Table \ref{tab:feb2006stat}). The statistics are calculated with hourly output over the three-day period displayed in Figure \ref{fig:feb2006}. To begin with, the mean absolute error in the water level ranges from 7.6 cm at Victoria to 13.1 cm at Campbell River, whereas, the mean absolute error in the residuals is smaller at all stations, however, this measure does not account for the large variability in water levels due to the tides. A statistic that does take into account this variability is $\gamma^2$ which suggests very good agreement in both the predicted water levels and residuals. Good agreement is also reflected in high Willmott skill scores at all locations. Note that the $\gamma^2$ measures and Willmott skill scores degrade slightly for the residuals, likely because the well-tuned tidal signal dominates the water level predictions.

@@

Next, an assessment of the importance of including the tide-surge interaction, which is important in shallow regions due to nonlinear effects and bottom friction, is presented for this hindcast (Figure \ref{fig:tidesurge}). The most significant differences occur at Point Atkinson where the maximum of the surge-only case is 8.7 cm higher than maximum modelled residual, about 10\% of the total surge amplitude.  In contrast, the tide-surge interaction at Victoria is hardly noticeable. Indeed, a spatial map of the difference between the modelled residual and surge-only difference (not shown here) indicates that the tide-surge interaction is more pronounced after the San Juan and Gulf Islands. The passages between these islands are known for vigorous tidal mixing, which may have a nonlinear effect on the surge propagation. However, the effect is a relatively small overestimate of the surge elevation and no changes to the timing of the surge. So, in operational settings, performing a surge-only simulation and adding the predicted tides \textit{a posteriori} could be more computationally efficient with only a small decrease in the accuracy of the prediction.

@@

An additional storm surge hindcast for Nov 2006 is provided as another means of assessing model performance and is mentioned briefly here. On Nov 16, 2006 communities in the Strait of Georgia experienced a strong wind storm with speeds upwards of 20 m s$^{-1}$ near Point Atkinson resulting in observed water levels of 5.1 m and a maximum surge amplitude of 83.7 cm. The surge amplitude was lower at the other stations, reaching no more than 68 cm. The model performs reasonably well in this hindcast; Willmott skill scores are greater than 0.94 (Table \ref{tab:nov2006stat}).  The best match in maximum surge occurred at Campbell River with a 6.2 cm underestimate by the model. The timing of the maximum surge at Campbell River was identical with observations and too early at the other the stations.

@@

Next, an assessment of the factors that are most important in scenarios leading up to storm surges is presented for the Feb 2006 storm surge (Figure \ref{fig:factors}).Four simulations were done: First, a simulation with all the forcing conditions, including tides, atmospheric forcing, rivers, and sea surface height anomaly. Second, a simulation with all forcing except the atmospheric forcing. Third, a simulation with all forcing except the sea surface height anomaly. Fourth, a simulation with only wind forcing, that is no anomaly and no inverse barometer effect due to pressure. It is clear that the sea surface height anomaly contributes most significantly to the surge at each of these locations (Figure \ref{fig:factors}). When the sea surface height anomaly is not included as a forcing condition, the surge drops from a maximum 82 cm to 11 cm at Point Atkinson and its maximum occurs eight hours later.  A similar drop in the maximum surge and delay in timing of the maximum surge is observed at the other locations. The atmospheric model indicates southeasterly winds near Point Atkinson during the onset of the surge but after the peak surge has passed the winds shifted to westerly. A strong westerly wind would cause an elevated sea surface height at Point Atkinson as water piles against the coast, explaining the delay in peak surge when only atmospheric forcing is used since the winds shift to westerly after the peak surge has passed. A simulation with only wind forcing produces a surge of 7.8 cm at Point Atkinson, a small fraction of the total observed surge, and only centimetres at the other locations. Note that this wind-induced surge occurs well after the peak surge occurred in both the observations and simulations with full forcing.

@@

These comparisons point to the importance of including, as a forcing condition at the open boundary, the surge entering the domain from the Pacific Ocean. Storms travelling over the Pacific Ocean towards the west coast of the northern United States and Canada can induce elevated sea levels along the coastline which then enter the Salish Sea system through the Strait of Juan de Fuca. In a model domain that does not include the Pacific Ocean, this effect must be included as a boundary condition. Previous simulations by \citet{murty1995storm} suggest that the inverse barometer effect and the surge entering the system through the Strait of Juan de Fuca contribute most significantly to storm surges in this domain.

@@

\subsection{Spatial Extent of Surge}
On Dec 15, 2006, a major storm with wind gusts over 120 km hr$^{-1}$ caused significant damage to many coastal communities on the Pacific Northwest. Although the storm was very strong and the observed residual at Point Atkinson reached 84 cm, this event did not coincide with high tide and so no significant or reported flooding occurred in the Vancouver area \citep{forseth2006adaptation}. Because of the high winds, this event has been used to examine the propagation of the surge throughout the domain and spatial variations in the surge amplitude. First, an assessment of the accuracy of the hindcast is provided.

Point Atkinson observed a maximum surge amplitude of 84 cm, which is approximately 28 cm higher than the modelled residual, however the timing of the surge agrees almost exactly as displayed in Figure \ref{fig:dec2006}. The best agreement occurs at Victoria where the observed surge was 62 cm and the model returned 44 cm. At all of the locations, the model predicted the maximum surge to occur within 3 hours of the observed surge. Typically, the modelled surge was too early, in contrast to the later predictions in the Feb 4, 2006 case. All Willmott skill scores rank above 0.89 and the $\gamma^2$ measure also indicates good agreement between model and observations (Table \ref{tab:dec2006stat}). The modelled residuals are on average too low and have a lower skill score than the Feb 2006 case. This could be due to a number of factors. First, the very large wind speeds observed during this storm are significantly underrepresented by the atmospheric model leading to lower wind stress and less piling of water against the coast. Second, the anomaly forcing at the open boundaries could be too low. The observed residual at Tofino, BC (a point near by but outside of the domain) is up to 10 cm lower than the observed residual at Neah Bay, WA (a point near by but inside of the domain). Since the storm surge is less sensitive to wind forcing at the surface than anomaly forcing at the open boundaries, the latter is likely the cause.

@@

In a simulation without atmospheric forcing, the surge amplitude across the Strait of Georgia is uniform with no pronounced difference between the mainland side and Vancouver Island side (Figure \ref{fig:spatial_sshonly}). While the surge entering the domain from the Pacific Ocean is the leading order contribution, the winds may induce localized changes to the surge amplitude. Interestingly, the winds over the Strait of Juan de Fuca cause a decrease in surge amplitude in this region as the winds are acting to push the water towards the mainland.  Even without atmospheric forcing, there is still a significant difference in surge amplitude between the Strait of Georgia and the Strait of Juan de Fuca, perhaps because the Strait of Georgia is also affected by the surge entering the domain from Johnstone Strait at the northern boundary or the incoming surge wave is partially reflected from the mainland coast.

@@

\subsection{Strong Wind Event}
Next, a relatively strong wind event recorded at Point Atkinson on Nov 19, 2009 is considered. As recorded in the Environment Canada Climate Data archives \citep{ECClimateArchive}, the maximum hourly wind speed at Point Atkinson was 18.6 m s$^{-1}$ and the largest gust observed was 25.8 m s$^{-1}$ or 93 km hr$^{-1}$. In addition, large wind speeds over 15 m s$^{-1}$ persisted for over 9 hours. Yet, the surge associated with this event is rather small when compared to the previous two cases (Figure \ref{fig:nov2009}). Two significant surge events are noted, the first occurring on Nov 19 with an observed amplitude of about 50 cm at Point Atkinson, and a second, larger surge of 62.5 cm on Nov 20. Surprisingly, the strong wind event of Nov 19 was associated with the smaller of the two surges. Neither of the surges occurred during a high tide and so no high water level events were recorded. Although not shown here, the other three locations also observed a maximum surge on Nov 20, which appears to be linked to a secondary high wind event and another surge entering the system at Tofino and Port Hardy.

\citet{danard2003storm} discuss the role of wind fetch, the distance over which wind acts on the surface of a water basin, in setting up seiche motion in lakes and its contribution to storm surges. As wind blows over the surface of a lake, the water level becomes elevated on the downwind side and depressed on the upwind side in response to the wind-induced currents. The longer the fetch, the more significant the tilting of the water surface. These changes in water level can also contribute to storm surges.  During the high wind events on Nov 19 and Nov 20, the winds at Point Atkinson were directed to the west, explaining the lack of strong surge observed during this event since wind stresses would act to depress the water on the east side of the basin. In fact, the surge amplitude at Victoria on Nov 20 was 64.7 cm which is slightly higher than that observed at Point Atkinson (62.5 cm). Of the cases considered so far, this is the only example where the surge amplitude at Victoria was higher than that at Point Atkinson. This event observed a higher surge at Victoria than the Dec 2006 event, further highlighting the role that winds plays in setting up local variations in the surge. In this case, the model results do not reflect a higher surge at Victoria (46.1 cm versus 50.5~cm at Point Atkinson), perhaps due to an underestimate in the wind speed and some discrepancies in the wind direction of the atmospheric forcing.
